Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-11-23 manuscript/manuscript.Rmd
minor updates.

  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change"), 
                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip.", "Rotation and/or reorientation of the core involving repositioning of the hand.", "Abrasion of core edges using a hammerstone", "Exchange of one percussor for another"), check.names=FALSE)

@@

The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure, which unfortunately still does not exist yet in experimental archaeology. Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

>>>>

  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change", "Winding Up"), 
                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip, which often occurs during the core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand, which is often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone. The abrasion movement should come from at least two different directions.", "Exchange of one percussor for another", "Preparational percussor movements towards the core that do not lead to the detachment of flakes, which can either be in direct contact with cores or not."), check.names=FALSE)

@@

The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2018], which unfortunately still does not exist yet in experimental archaeology. The accessibility and availability of experimental data can foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data. Moreover, a centralized database also promotes data preservation and long-term accessibility. By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time. This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments. It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated. However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].

Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project. Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researcher; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
